Optimal management of a stochastically varying population when policy adjustment is costly

### 2015-02-03 manuscripts/manuscript.Rmd
fine tuning figures

---
title: Optimal management of a stochastically varying population when policy adjustment is costly
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig (at) gmail.com
    footnote: Corresponding author
  - name: Michael Bode
    affiliation: melbourne
  - name: James Sanchirico
    affiliation: UCD
  - name: Jacob LaRiviere
    affiliation: UTKEcon
  - name: Alan Hastings
    affiliation: UCD
  - name: Paul R. Armsworth 
    affiliation: UTKEEB

@@

To illustrate this tradeoff, we use an example from fisheries
management. Figure 1, top-left panel, shows a time series for the
estimated population size, represented as spawning stock biomass, of
west Atlantic bluefin tuna (_Thunnus thynnus_, henceforth bluefin) from
a recent stock assessment [@ICCAT2009a].  The bottom-left panel 
shows the catch quota (Total Allowable Catch or TAC, before being adjusted
for overages or underages in the catch of each country in the previous year)
for the stock as set by the relevant
management agency [@ICCAT2009a]. Despite the estimated population size
declining by `r 1 - min(tuna$stock)/max(tuna$stock)`% between 1982 and 1991, the quota was not changed during
this period. Instead, the quota only changed occasionally and in between
times was left unaltered. Fishery management decisions regarding this
species can be highly contentious (@Safina1998, @Sissenwine1998,
@Porch2005). Moreover, the stock is fished by fleets from many nations
with quotas being set by a multilaterial management agency through a
process of negotiation. As such, we might reasonably anticipate that
for this species there could be substantial transaction costs involved
in reaching agreement over any quota change, or constraints on the frequency
with which the those changes could occur.  Either could contribute to
the observed quota stability.

@@

As the examples in Figure 1 make clear, management policies will rarely be
as responsive as this constant escapement policy assumes. In this paper,
we explore conditions under which this lack of responsiveness may be a
rational response on the part of managers. Specifically, we consider a
case where managers seek to balance the benefits in terms of increased
profits from fishing from more finely tracking recruitment variations
with the growing costs associated with adjusting policies frequently to do
so. The policy adjustment costs involved could reflect pure administrative
transaction costs or preferences held by fishermen, fish processing plants
or other stakeholders, for less variable quotas. In seeking to account
for these policy adjustment costs, we recognize that we do not know just
what functional form they should take and that it will require a lot of
empirical work to estimate that. Therefore, we scope three candidate
functional forms that represent qualitatively different assumptions about how these
costs operate to determine whether the results we obtain are sensitive
to such differences.

@@

To simplify presentation of the results, we will illustrate cases where the
growth parameters are chosen so that the intrinsic rate of increase of
the fish population is 1 and the equilibrium biomass for the equivalent
deterministic model without harvesting is `r K`; (specifically, $A=$ `r pars[1]` and
$B=$ `r pars[2]` in Eqn \eqref{eq:state_equation}). To characterize environmental variability, we
assume multiplicative shock $Z_t$ is distributed log-normally with
log standard deviation $\sigma_g$ = `r sigma_g`. In addition, we
show cases where $p=$ `r price`, $\delta=$ `r delta` and $c_0=$ `r c0`.
In the supplement we illustrate that the qualitative patterns observed
here are not sensitive to the specific choices of these parameter values.

@@

ggplot(fees, aes(c2, (npv0-value)/npv0, lty=variable, color=variable)) + 
  geom_line() + 
  geom_hline(aes(yintercept=reduction), linetype=4) + 
  xlab("Penalty coefficient") +
  ylab("Reduction in net present value") + 
  scale_linetype_discrete(labels=relabel, breaks = c('L1', 'L2', 'fixed')) + 
  scale_color_discrete(labels=relabel, breaks = c('L1', 'L2', 'fixed')) 
 # rename scale from 'variable' to 'adjustment cost for policy:'

@@

@@

```{r Figure_4, fig.cap="Variance and autocorrelation of harvest dynamics, as a function of the penalty coefficient.  Adjustment penalty is calculated as percentage of the maximum expected NPV in the basic case with no policy adjustment costs.", fig.height=6, fig.width=4, cache=fig.cache, dev=c("pdf", "svg", "png")}

@@

p4 <- ggplot(f4df, aes(x = penalty_fraction, fill=penalty_fn, col=penalty_fn)) + 
  coord_cartesian(xlim = c(0, .3)) + 
  scale_color_discrete(labels=relabel, name="Adj. cost model") + 
  scale_fill_discrete(labels=relabel, name="Adj. cost model")

@@

legend <- gtable_filter(ggplot_gtable(ggplot_build(p4a)), "guide-box") 
grid.arrange(do.call(arrangeGrob, 
            lapply(list(p4a, p4b),
                   function(x) x + theme_bw() + theme(legend.position="none"))),
    legend,
    ncol = 2,
    widths = c(5,1))

@@

While still not common for the particular parameter combinations we examine,
We find that positive unaltered quotas through time are much more likely
to occur when optimizing against $\Pi_1$ where policy adjustment costs
scale linearly with the size of proposed policy changes: over the 100 replicates
time series the harvest policy is strictly positive and identical in consecutive intervals
only `r no_shifts[["L2"]] * 100`% of the time for quadratic costs $\Pi_2$, compared with
`r no_shifts[["L1"]] * 100`% with linear costs $\Pi_1$ and `r no_shifts[["fixed"]]*100`%
for fixed costs $\Pi_3$. Moreover, these occurrences increase in frequency as the severity of
these costs ($c_i$) increases. Note that the restriction of only positive quotas lets us 
distinguish between cases that are constant due purely to adjustment costs from cases that
are constant purely due to boundary effects.

@@

fig5_dat <- hist_dat %>% 
  filter(penalty_fn=="L1") %>%
  select(-penalty_fn) %>%
  unite(dist, Assumption, variable, sep = " : ")

ggplot(fig5_dat) + 
  geom_density(aes(value, fill=dist, color=dist), alpha=0.8)

@@

ggplot(filter(table1_long, reduction == 0.25),
  aes(penalty.fn, value, fill = variable)) + 
  geom_bar(stat="identity", position="dodge") + 
  facet_wrap(~reduction, ncol=2) + 
  scale_x_discrete(labels = c('L1' = substitute(paste(Pi[1])),
                              'L2' = substitute(paste(Pi[2])),
                              'fixed' = substitute(paste(Pi[3])))) + 
  xlab("Penalty function") + 
  ylab(expression(NPV[0]))
write.csv(table1_long, "components/data/figure6.csv")
```

@@

The biggest differences between the representations of policy adjustment
cost that we consider are between those that have a smoothing effect on
annual quotas and those that do not. The few past studies that incorporate
policy adjustment costs in models of fisheries or other environmental
management contexts [@Feichtinger1994; @Wirl1999] have assumed costs of
policy adjustment increased as a quadratic function of the magnitude of
the policy change being proposed (analogous to $\Pi_2$ formulation). In
effect, this means that small changes to annual quotas incur little
extra cost, but large changes to annual quotas become disproportionately
expensive to make. Including costs of this form smooths inter-annual
variation in the recommended catch quotas. Also, the catches
and remaining stock sizes corresponding to optimal management become more
autocorrelated in time, because it takes the fishery multiple timesteps to harvest
down peaks in abundance that follow large recruitment pulses. The effects
of smoothing here are similar to those predicted when assuming the cost
per unit effort is increasing in the amount of effort expended (e.g.,
@Lewis1981, @McGough2009; see Supporting Information for a summary of the
relevant results) as opposed to associating extra costs with changes to
policy *per se*. Smoothing effects of this type are what one might have
expected when including policy adjustment costs.

@@

That the different representations of policy adjustment costs result in
such different dynamics suggests that researchers constructing fisheries
economics models should proceed cautiously when choosing how to represent
these costs. However, in our own experience, we have found that while
modeling studies sometimes mention policy adjustment costs when motivating
model assumptions [@Lewis1981; @Feichtinger1994; @Wirl1999; @McGough2009],
they rarely provide much justification
for the choice of functional form used or test the sensitivity of any
conclusions drawn to alternative specifications. Indeed, we were surprised
to find such clear differences between the functional forms, because
we anticipated that the limiting case of increasing policy adjustment
costs within each functional form should be the same, namely a constant
annual quota that does not change through time (open loop control).

@@

As with any modeling analysis, our formulation makes many assumptions. For
example, we assume that the fishery in question is being optimally managed
by a policymaker who acts as the "sole owner" of the stock. This approach
is different to models that assume fisheries are not well managed,
e.g. by assuming open access conditions,
or models that derive policy recommendations endogenously by modeling
interactions between different stakeholders [@Kaitala1993; @Laukkanen2003].
As such, we anticipate that our approach will be more relevant to
some fisheries, particularly domestic fisheries in developed countries
that are subject to relatively strong regulatory regimes, than to others
(e.g., artisanal fisheries that are subject to weaker regulation). To
focus on the effects of policy adjustment costs, we focused narrowly
on the basic model specification of @Reed1979 and examined how the
predictions of this classic problem were changed by introducing policy
adjustment costs of different forms. However, there have been many
elaborations on Reed's basic approach that increase the realism of
the optimization models involved by relaxing other assumptions (see for
example @Sethi2005; @Singh2006; @McGough2009).

@@

This work was assisted through participation in "Pretty Darn Good Control"
Investigative Workshop at the National Institute for Mathematical and
Biological Synthesis, sponsored by the National Science Foundation through
NSF Award #DBI-1300426, with additional support from The University of
Tennessee, Knoxville.  The authors acknowledge helpful discussions and
input from working group co-organizers Megan Donahue and Carl Toews, and
participants Marie-Josee Fortin, Dan Ryan, Frank Doyle, Claire Paris,
Iadine Chades and Mandy Karnauskas.  We also acknowledge the following
support: NSF Grant DBI-1306697 (CB), ARC DECRA (DE130100572) and The
ARC Centre of Excellence for Environmental Decisions (MB), ...

>>>>

---
title: Optimal management of a stochastically varying population when policy adjustment is costly
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig (at) gmail.com
    footnote: Corresponding author
  - name: Michael Bode
    affiliation: melbourne
  - name: James N. Sanchirico
    affiliation: UCD
  - name: Jacob LaRiviere
    affiliation: UTKEcon
  - name: Alan Hastings
    affiliation: UCD
  - name: Paul R. Armsworth 
    affiliation: UTKEEB

@@

To illustrate this tradeoff, we use an example from fisheries
management. Figure 1, top-left panel, shows a time series for the
estimated population size, represented as spawning stock biomass, of
west Atlantic bluefin tuna (_Thunnus thynnus_, henceforth bluefin) from
a recent stock assessment [@ICCAT2009a].  The bottom-left panel 
shows the catch quota (Total Allowable Catch or TAC, before being adjusted
for overages or underages in the catch of each country in the previous year)
for the stock as set by the relevant
management agency [@ICCAT2009a]. Despite the estimated population size
declining by `r 100 * (1 - min(tuna$stock) / max(tuna$stock))`% between
1982 and 1991, the quota was not changed during
this period. Instead, the quota only changed occasionally and in between
times was left unaltered. Fishery management decisions regarding this
species can be highly contentious (@Safina1998, @Sissenwine1998,
@Porch2005). Moreover, the stock is fished by fleets from many nations
with quotas being set by a multilaterial management agency through a
process of negotiation. As such, we might reasonably anticipate that
for this species there could be substantial transaction costs involved
in reaching agreement over any quota change, or constraints on the frequency
with which the those changes could occur.  Either could contribute to
the observed quota stability.

@@

As the examples in Figure 1 make clear, management policies will rarely be
as responsive as this constant escapement policy assumes. In this paper, we consider a
case where managers seek to balance the benefits in terms of increased
profits from fishing from more finely tracking recruitment variations
with the growing costs associated with adjusting policies frequently to do
so. The policy adjustment costs involved could reflect pure administrative
transaction costs or preferences held by fishermen, fish processing plants
or other stakeholders, for less variable quotas. In seeking to account
for these policy adjustment costs, we recognize that we do not know just
what functional form they should take and that it will require substantial 
empirical work to estimate that. Therefore, we investigate three candidate
functional forms that represent qualitatively different assumptions about how these
costs operate to determine whether the results we obtain are sensitive
to such differences.

@@

To simplify presentation of the results, we will illustrate cases where the
growth parameters are chosen so that the equilibrium biomass for the equivalent
deterministic model without harvesting is `r K`; (specifically, $A=$ `r pars[1]` and
$B=$ `r pars[2]` in Eqn \eqref{eq:state_equation}). To characterize environmental variability, we
assume multiplicative shock $Z_t$ is distributed log-normally with
log standard deviation $\sigma_g$ = `r sigma_g`. In addition, we
show cases where $p=$ `r price`, $\delta=$ `r delta` and $c_0=$ `r c0`.
In the supplement we illustrate that the qualitative patterns observed
here are not sensitive to the specific choices of these parameter values.

@@

ggplot(fees, aes(c2, (npv0-value)/npv0, lty=variable, color=variable)) + 
  geom_line() + 
  geom_hline(aes(yintercept=reduction), linetype=4) + 
  xlab("Penalty coefficient") +
  ylab("Reduction in net present value") + 
  scale_linetype_discrete(labels=relabel, breaks = c('L1', 'L2', 'fixed'), name = 'cost model') + 
  scale_color_discrete(labels=relabel, breaks = c('L1', 'L2', 'fixed'), name = 'cost model') 
 # rename scale from 'variable' to 'adjustment cost for policy:'

@@

dt3$penalty_fn <- factor(dt3$penalty_fn, levels=c("L1", "L2", "fixed"))

@@

```{r Figure_4, fig.cap="Variance and autocorrelation of harvest dynamics, as a function of the penalty coefficient.  Adjustment penalty is calculated as percentage of the maximum expected NPV in the basic case with no policy adjustment costs.", fig.height=6, fig.width=5, cache=fig.cache, dev=c("pdf", "svg", "png")}

@@

p4 <- ggplot(f4df, aes(x = penalty_fraction, fill=penalty_fn, col=penalty_fn)) + 
  coord_cartesian(xlim = c(0, .3)) + 
  scale_color_discrete(labels=relabel, name="cost model") + 
  scale_fill_discrete(labels=relabel, name="cost model")

@@

legend <- gtable_filter(ggplot_gtable(ggplot_build(p4a)), "guide-box") 
grid.arrange(do.call(arrangeGrob, 
            lapply(list(p4a, p4b),
                   function(x) x + theme_bw() + theme(legend.position="none"))),
    legend,
    ncol = 2,
    widths = c(5,2))

@@

While still not common for the particular parameter combinations we examine,
we find that positive unaltered quotas through time are much more likely
to occur when optimizing against $\Pi_1$ where policy adjustment costs
scale linearly with the size of proposed policy changes: over the 100 replicates
time series the harvest policy is strictly positive and identical in consecutive intervals
only `r no_shifts[["L2"]] * 100`% of the time for quadratic costs $\Pi_2$, compared with
`r no_shifts[["L1"]] * 100`% with linear costs $\Pi_1$ and `r no_shifts[["fixed"]]*100`%
for fixed costs $\Pi_3$. Moreover, these occurrences increase in frequency as the severity of
these costs ($c_i$) increases. Note that the restriction of only positive quotas lets us 
distinguish between cases that are constant due purely to adjustment costs from cases that
are constant purely due to boundary effects.

@@

fig5_dat <- hist_dat %>% 
  filter(penalty_fn=="L1") %>%
  select(-penalty_fn) %>%
  unite(dist, Assumption, variable, sep = " : ")
labels <- c(expression(paste("Costs under ",  Pi[1])), 
            expression(paste("Revenue under ", Pi[1])),
            expression(paste("Cost under ", Pi[0])), 
            expression(paste("Revenue under ", Pi[0])))
ggplot(fig5_dat) + 
  geom_density(aes(value, fill=dist, color=dist, lty=dist), alpha=0.5, size=1) +
  scale_fill_discrete(name="", labels=labels) + 
  scale_color_discrete(labels=labels, name="") + 
   scale_linetype_discrete(labels=labels, name="") +
  xlab("Realized NPV")

@@

ggplot(filter(table1_long, reduction == 0.25),
  aes(penalty.fn, value, fill = variable)) + 
  geom_bar(stat="identity", position="dodge") + 
  scale_fill_discrete(name="error type") + 
  scale_x_discrete(labels = c('L1' = substitute(paste(Pi[1])),
                              'L2' = substitute(paste(Pi[2])),
                              'fixed' = substitute(paste(Pi[3])))) + 
  xlab("Cost model") + 
  ylab(expression(NPV[0])) + theme_tufte(base_size=16)
write.csv(table1_long, "components/data/figure6.csv")
```

@@

The biggest differences between the representations of policy adjustment
cost that we consider are between those that have a smoothing effect on
annual quotas and those that do not. The few past studies that incorporate
policy adjustment costs in models of fisheries or other environmental
management contexts [@Feichtinger1994; @Wirl1999] have assumed costs of
policy adjustment increased as a quadratic function of the magnitude of
the policy change being proposed (analogous to $\Pi_2$ formulation). In
effect, this means that small changes to annual quotas incur little
extra cost, but large changes to annual quotas become disproportionately
expensive to make. Including costs of this form smooths inter-annual
variation in the recommended catch quotas. Also, the catches
and remaining stock sizes corresponding to optimal management become more
autocorrelated in time, because it takes the fishery multiple timesteps to harvest
down peaks in abundance that follow large recruitment pulses. The effects
of smoothing here are similar to those predicted when assuming the cost
per unit effort is increasing in the amount of effort expended [@Brown1974;
@Lewis1981; @McGough2009]; (see Supporting Information for a summary of the
relevant results) as opposed to associating extra costs with changes to
policy *per se*. Smoothing effects of this type are what one might have
expected when including policy adjustment costs.

@@

That the different representations of policy adjustment costs result in
such different dynamics suggests that researchers constructing fisheries
economics models should proceed cautiously when choosing how to represent
these costs. However, in our own experience, we have found that while
modeling studies sometimes mention policy adjustment costs when motivating
model assumptions [@Brown1974; @Lewis1981; @Feichtinger1994; @Wirl1999; @McGough2009],
they rarely provide much justification
for the choice of functional form used or test the sensitivity of any
conclusions drawn to alternative specifications. Indeed, we were surprised
to find such clear differences between the functional forms, because
we anticipated that the limiting case of increasing policy adjustment
costs within each functional form should be the same, namely a constant
annual quota that does not change through time.

@@

As with any modeling analysis, our formulation makes many assumptions. For
example, we assume that the fishery in question is being optimally managed
by a policymaker who acts as the "sole owner" of the stock. This approach
is different to models that assume fisheries are not well managed,
e.g. by assuming regulated open access conditions [@Homans1997],
or models that derive policy recommendations endogenously by modeling strategic 
interactions between different stakeholders [@Kaitala1993; @Laukkanen2003].
As such, we anticipate that our approach will be more relevant to
some fisheries, particularly domestic fisheries in developed countries
that are subject to relatively strong regulatory regimes, than to others
(e.g., artisanal fisheries that are subject to weaker regulation). To
focus on the effects of policy adjustment costs, we focused narrowly
on the basic model specification of @Reed1979 and examined how the
predictions of this classic problem were changed by introducing policy
adjustment costs of different forms. However, there have been many
elaborations on Reed's basic approach that increase the realism of
the optimization models involved by relaxing other assumptions (see for
example @Sethi2005; @Singh2006; @McGough2009).

@@

This work was assisted through participation in "Pretty Darn Good Control"
Investigative Workshop at the National Institute for Mathematical and
Biological Synthesis, sponsored by the National Science Foundation through
NSF Award #DBI-1300426, with additional support from The University of
Tennessee, Knoxville.  The authors acknowledge helpful discussions and
input from working group co-organizers Megan Donahue and Carl Toews, and
participants Marie-Josee Fortin, Dan Ryan, Frank Doyle, Claire Paris,
Iadine Chades and Mandy Karnauskas.  We also acknowledge the support of
NSF Grant DBI-1306697 to CB and ARC DECRA (DE130100572) & the
ARC Centre of Excellence for Environmental Decisions to MB.
